Structural replication of nanoporous media using procedural noise

### 2019-01-03 ms.tex
Fixes

\documentclass[aps,pre,twocolumn,letterpaper,floatfix,showpacs]{revtex4}
\usepackage{graphicx} 
\usepackage{amsmath,amssymb,amsfonts} 
\usepackage{mathtools}
\usepackage{pdfpages}
\usepackage{afterpage}
\usepackage[hidelinks]{hyperref} 
\usepackage{epstopdf}
\usepackage{siunitx}
\usepackage{color}
\definecolor{Blue}{rgb}{0.1,0.1,1.0} 
\definecolor{Magenta}{rgb}{1.0,0.1,0.5} 
\definecolor{LRed}{rgb}{0.8,0.0,0.0}

 \newcommand{\todo}[1]{ {\color{Magenta} TODO: \color{Blue} \textbf{#1} }}

\begin{document}
\title{Structural replication of nanoporous media using procedural noise}
%\author{A. Hafreager, N. Groeneboom, A. Malthe-S\oe renssen$^1$}
\email{anders.hafreager@fys.uio.no}
\author{Anders Hafreager $^{1}$} 
\author{Nicolaas Groeneboom $^{1}$} 
\author{Anders Malthe-S\o renssen $^1$}
\affiliation{$^1$Department of Physics - University of Oslo\\Sem S{\ae}lands vei 24, NO-0316, Oslo, Norway }
\date{\today}

@@

\section{Procedural content generation}
By procedural content generation, we generally mean a mathematical algorithm that uses limited
input data to create complex, sometimes infinite repeating patterns.
A firm example would be fractals, such as the Julia/Mandelbrot [ref Mandelbrot] set, where one
tests whether the complex quadratic recurrence equation $z_{n+1} = z_n^2 +C$ for $z \in \mathbb C$
converges for each position in a 2D complex plane.
With this seemingly simple equation, the level of complexity generated on the boundary between
convergence and divergence is infinite, with self-similar patterns repeating themselves to arbitrary detail.

@@

A much simpler example would be generating a 1 (or 2)-dimensional terrain by choosing random
points on a grid and performing a spline interpolation between them to create the illusion of a smooth landscape.
In fact, this is the very essence of procedural noise, which we are focusing on in our model,
and the first person to create such a fast algorithm for N-dimensional spline interpolation
between pseudo-random points was Ken Perlin\cite{perlin1985image}.

@@

\subsection{Properties of Perlin/simplex noise}
\begin{figure}
\includegraphics[width=.5\textwidth]{noise.eps}
\caption{Perlin noise (left) versus simplex noise (right). Note the
  straight x-y pattern in the single upper Perlin mode and the less
  anisotropic simplex mode. Lower part: combining octaves tend
  to remove these patterns.}
\label{fig:noise}
\end{figure}

\begin{figure}
\includegraphics[width=.5\textwidth]{noise_comparison.png}
%\mbox{\epsfig{file=perlinstats.eps, width=\linewidth}}
\caption{Normalized distributions of Perlin noise and simplex noise compared with a
  Gaussian distribution. Note how both noise functions produce
  near-Gaussian, but slightly different distributions.}
\label{fig:properties}
\end{figure}
Standard Perlin noise is known for having several anisotropic features
\citep{lagae:2010}. Perlin noise also gives rise to a ``textile''
pattern in the $x$ and $y$-directions, which is evident from the upper
left part of figure \ref{fig:noise}, depicting 2D Perlin noise in
a single octave. The upper-right part shows the same node for simplex noise,
with less obvious ``textile'' features. However, when combining several octaves
and adding a random shift to the coordinates, as depicted in the two lower parts,
these anisotropic patterns tend to cancel
out.  In addition, we have calculated the statistical distributions for Perlin and simplex noise
with $\sigma = 0.35$, and have plotted the results in figure \ref{fig:properties}.
Each procedural noise method has distributions that are close to normal, 
but with some asymmetrical properties.

@@

\begin{figure}
\includegraphics[width=.5\textwidth]{1d_noise.pdf}
\caption{1-dimensional simplex noise with 1 (yellow), 3 (red) and 30 (blue) octaves. Adding more octaves adds high frequency features, but it still follows the 1 octave base signal.}
\label{fig:1dperlin}
\end{figure}

@@

\begin{figure*}
\includegraphics[width=.95\textwidth]{model_examples.png}
\caption{Examples of 3D simplex noise structures in nanoporous media. Left image:
low persistence (large scale dominates), middle image: medium persistence (small
scales starts dominating) and right image: high threshold (more atoms are discarded). }
\label{fig:model_example}
\end{figure*}

@@

The posterior probability distribution is defined as the probability of the parameters
$\bar \theta$ given the evidence $d$, or $P(\bar \theta | d)$. Bayes' theorem states that 
\begin{equation}
P(\bar \theta | d) = \frac{P(d | \bar \theta)P(\bar \theta)}{P(d)} \propto P(d | \bar \theta)P(\bar \theta),
\end{equation}
where $P(d | \bar \theta)$ is called the Likelihood $\mathcal L$ (the probability of
evidence given parameters) and $P(\bar \theta)$ is the prior (the beliefs about the
quantity before some evidence is taken into account). $P(d)$ is called the evidence,
and is the probability of the observations (here assumed to be 1). 
To calculate the posterior probability distribution and obtain the best fit model parameters, we need only
to estimate the likelihood function, which is connected to the $\chi^2$\todo{Why do we have an exponential dependence of $\chi^2$ here?} by 
\begin{equation}
\mathcal L \propto e^{-\chi^2}.
\end{equation}

@@

\section{Results}
We now turn our attention to applying the MCMC framework on two kinds of data.
First, we wish to test and verify the code by performing a full data analysis on
mock data with known input parameters. This enables us to investigate the posterior
distribution, and determine whether any potential problems could arise with the
likelihood landscape, such as random walkers trapped in local minima. In addition,
we will obtain an intuition on the width of the estimated parameter distributions.
But most importantly, we will have verified that our likelihood code is working correctly
if we successfully reproduce the original input parameters. We then continue with a
full analysis of physical system simulated with molecular dynamics, and seek parameters that reproduces $g(r)$ of this system.
\todo{Figure 5 is the same as left in fig 9.}
\subsection{Model verification: Parameter estimation of mock data}
\begin{figure}[htb!]
\includegraphics[width=.45\textwidth]{model_test.png}
\caption{Simulated mock data used for the initial model testing analysis. Input parameters are $\nu=0.03$,$\tau=-0.2$ and $\psi=0.6$.}
\label{fig:mockdata}
\end{figure}
In order to verify our likelihood code, we perform a model verification test
on a mock data set with known input values based on equation \ref{eq:noisemodel1}.
In this example, we only use standard simplex noise with three free parameters:
the persistence $\psi$, threshold $\tau$ and scale $\nu$.

@@

\begin{figure*}
\includegraphics[width=.99\textwidth]{mock_data_results_pts.png}
\caption{Results from the model validation simulation.The upper panel displays the marginalised 1D posteriors for each parameter, while the lower panel depicts the marginalised 2D posteriors. Note that all input parameters were reproduced well within $1 \sigma$ (inner contour line).}
\label{fig:mockdataresults_pts}
\end{figure*}

\subsection{Parameter estimation of simulated nanoporous silica}
%\begin{figure}
%\includegraphics[width=.49\textwidth]{sio_porous.png}
%\caption{Simulated silica created with LAMMPS. The box is $159^3 \AA$, with $100 000$ particles, having been created by cooling an expanding system until quenching was achieved. Note the similarities between the pores here and in figure \ref{fig:mockdata}.}
%\label{fig:si2data}
%\end{figure}
\begin{figure*}
\includegraphics[width=.99\textwidth]{results_porous_full.png}
\caption{
Results from the analysis of a simulated $\SI{159} {\angstrom}^3$ SiO$_2$ nanoporous data set. The upper panel displays the marginalised 1D posteriors for each parameter, while the lower panel depicts the marginalised 2D posteriors. Best fit parameters from the maximum 3D likelihood suggest $\tau=-0.2$, $\psi=0.9$ and $\nu=0.03$.
}

@@

\begin{figure}
\includegraphics[width=0.5\textwidth]{gofr_figure.pdf}
\caption{$g(r)$ for three data sets: the simulated SiO$_2$ input data set (red), the best-fit parameter model (blue) and a random low-fit parameter model (yellow).
The red shaded area for low $r$ represents the part of $g(r)$ we have not used to calculate $\chi^2$ since it only captures nearest neighbor features, not the pore structure.
The main difference between the bad fit and the others is the scaling which is due to slow convergence since the bad fit dataset has very large pores.
There are two weak bumps at \SI{6.5}{\angstrom} and \SI{7.5}{\angstrom} which we don't pick up with our model, but the general behaviour is reproduced by our model both on short and long length scales.}
\label{fig:gofr1}
\end{figure}

Finally, we depict a comparison of the actual simulated input SiO$_2$ data set together with a
representation of the best-fit parameters in figure \ref{fig:porous_vs_model}   
\begin{figure}
\includegraphics[width=.45\textwidth]{comparison.png}
\caption{Comparison of our best fit model (left) with parameters $\tau=-0.2$, $\psi=0.9$ and $\nu=0.03$ and the simulated SiO$_2$ input data (right).
Although the pore geometry seems to be similar, we see some artifacts, such as single loose atoms in the generated geometry.
To obtain a structure that can be used further, one would have to passivate the system by adding atoms so each atom has the desired coordination number.
Both the porosity and the surface area of the model and the data are found to be similar.}
\label{fig:porous_vs_model}
\end{figure}

@@

\todo{skriv om hvorfor vi ikke plukker opp alle features? Men i diskusjon!}

>>>>

\documentclass[aps,pre,twocolumn,letterpaper,floatfix,showpacs]{revtex4}
\usepackage{graphicx} 
\usepackage{amsmath,amssymb,amsfonts} 
\usepackage{mathtools}
\usepackage{pdfpages}
\usepackage{afterpage}
\usepackage[hidelinks]{hyperref} 
\usepackage{epstopdf}
\usepackage{siunitx}
\usepackage{color}

\begin{document}
\title{Structural replication of nanoporous media using procedural noise}
\email{anders.hafreager@fys.uio.no}
\author{Anders Hafreager $^{1}$} 
\author{Nicolaas Groeneboom $^{1}$} 
\author{Anders Malthe-S\o renssen $^1$}
\affiliation{$^1$Department of Physics - University of Oslo\\Sem S{\ae}lands vei 24, NO-0316, Oslo, Norway }
\date{\today}

@@

\section{Procedural content generation}
By procedural content generation, we generally mean a mathematical algorithm that uses limited
input data to create complex, sometimes infinite repeating patterns.
A firm example would be fractals, such as the Mandelbrot set, where one
tests whether the complex quadratic recurrence equation $z_{n+1} = z_n^2 +C$ for $z \in \mathbb C$
converges for each position in a 2D complex plane.
With this seemingly simple equation, the level of complexity generated on the boundary between
convergence and divergence is infinite, with self-similar patterns repeating themselves to arbitrary detail.

@@

A much simpler example would be generating a 1 (or 2)-dimensional terrain by choosing random
points on a grid and performing a spline interpolation between them to create the illusion of a smooth landscape.
In fact, this is the very essence of procedural noise, which we are focusing on in our model,
and the first person to create such a fast algorithm for $N$-dimensional spline interpolation
between pseudo-random points was Ken Perlin\cite{perlin1985image}.

@@

\subsection{Properties of Perlin/simplex noise}
\begin{figure}
\includegraphics[width=.5\textwidth]{fig1.eps}
\caption{Perlin noise (left) versus simplex noise (right). Note the
  straight x-y pattern in the single upper Perlin mode and the less
  anisotropic simplex mode. Lower part: combining octaves tend
  to remove these patterns.}
\label{fig:noise}
\end{figure}

\begin{figure}
\includegraphics[width=.5\textwidth]{fig2.png}
%\mbox{\epsfig{file=perlinstats.eps, width=\linewidth}}
\caption{Normalized distributions of Perlin noise and simplex noise compared with a
  Gaussian distribution. Note how both noise functions produce
  near-Gaussian, but slightly different distributions.}
\label{fig:properties}
\end{figure}
Standard Perlin noise is known for having several anisotropic features
\citep{lagae:2010}. Perlin noise also gives rise to a ``textile''
pattern in the $x$ and $y$-directions, which is evident from the upper
left part of figure \ref{fig:noise}, depicting 2D Perlin noise in
a single octave. The upper-right part shows the same node for simplex noise,
with less obvious ``textile'' features. However, when combining several octaves
and adding a random shift to the coordinates, as depicted in the two lower parts,
these anisotropic patterns tend to cancel
out.  In addition, we have calculated the statistical distributions for Perlin and simplex noise
with $\sigma = 0.35$, and have plotted the results in figure \ref{fig:properties}.
Each procedural noise method has distributions that are close to normal, 
but with some asymmetrical properties.

@@

\begin{figure}
\includegraphics[width=.5\textwidth]{fig3.pdf}
\caption{1-dimensional simplex noise with 1 (yellow), 3 (red) and 30 (blue) octaves. Adding more octaves adds high frequency features, but it still follows the 1 octave base signal.}
\label{fig:1dperlin}
\end{figure}

@@

\begin{figure*}
\includegraphics[width=.95\textwidth]{fig4.png}
\caption{Examples of 3D simplex noise structures in nanoporous media. Left image:
low persistence (large scale dominates), middle image: medium persistence (small
scales starts dominating) and right image: high threshold (more atoms are discarded). }
\label{fig:model_example}
\end{figure*}

@@

The posterior probability distribution is defined as the probability of the parameters
$\bar \theta$ given the evidence $d$, or $P(\bar \theta | d)$. Bayes' theorem states that 
\begin{equation}
P(\bar \theta | d) = \frac{P(d | \bar \theta)P(\bar \theta)}{P(d)} \propto P(d | \bar \theta)P(\bar \theta),
\end{equation}
where $P(d | \bar \theta)$ is called the Likelihood $\mathcal L$ (the probability of
evidence given parameters) and $P(\bar \theta)$ is the prior (the beliefs about the
quantity before some evidence is taken into account). $P(d)$ is called the evidence,
and is the probability of the observations (here assumed to be 1). 
To calculate the posterior probability distribution and obtain the best fit model parameters, we need only
to estimate the likelihood function, which is connected to the $\chi^2$ by 
\begin{equation}
\mathcal L \propto e^{-\chi^2}.
\end{equation}

@@

\section{Results}
We now turn our attention to applying the MCMC framework on two kinds of data.
First, we wish to test and verify the code by performing a full data analysis on
mock data with known input parameters. This enables us to investigate the posterior
distribution, and determine whether any potential problems could arise with the
likelihood landscape, such as random walkers trapped in local minima. In addition,
we will obtain an intuition on the width of the estimated parameter distributions.
But most importantly, we will have verified that our likelihood code is working correctly
if we successfully reproduce the original input parameters. We then continue with a
full analysis of physical system simulated with molecular dynamics, and seek parameters that reproduces $g(r)$ of this system.
\subsection{Model verification: Parameter estimation of mock data}
\begin{figure}[htb!]
\includegraphics[width=.45\textwidth]{fig5.png}
\caption{Simulated mock data used for the initial model testing analysis. Input parameters are $\nu=0.03$,$\tau=-0.2$ and $\psi=0.6$.}
\label{fig:mockdata}
\end{figure}
In order to verify our likelihood code, we perform a model verification test
on a mock data set with known input values based on equation \ref{eq:noisemodel1}.
In this example, we only use standard simplex noise with three free parameters:
the persistence $\psi$, threshold $\tau$ and scale $\nu$.

@@

\begin{figure*}
\includegraphics[width=.99\textwidth]{fig6.png}
\caption{Results from the model validation simulation.The upper panel displays the marginalised 1D posteriors for each parameter, while the lower panel depicts the marginalised 2D posteriors. Note that all input parameters were reproduced well within $1 \sigma$ (inner contour line).}
\label{fig:mockdataresults_pts}
\end{figure*}

\subsection{Parameter estimation of simulated nanoporous silica}
%\begin{figure}
%\includegraphics[width=.49\textwidth]{sio_porous.png}
%\caption{Simulated silica created with LAMMPS. The box is $159^3 \AA$, with $100 000$ particles, having been created by cooling an expanding system until quenching was achieved. Note the similarities between the pores here and in figure \ref{fig:mockdata}.}
%\label{fig:si2data}
%\end{figure}
\begin{figure*}
\includegraphics[width=.99\textwidth]{fig7.png}
\caption{
Results from the analysis of a simulated $\SI{159} {\angstrom}^3$ SiO$_2$ nanoporous data set. The upper panel displays the marginalised 1D posteriors for each parameter, while the lower panel depicts the marginalised 2D posteriors. Best fit parameters from the maximum 3D likelihood suggest $\tau=-0.2$, $\psi=0.9$ and $\nu=0.03$.
}

@@

\begin{figure}
\includegraphics[width=0.5\textwidth]{fig8.pdf}
\caption{$g(r)$ for three data sets: the simulated SiO$_2$ input data set (red), the best-fit parameter model (blue) and a random low-fit parameter model (yellow).
The red shaded area for low $r$ represents the part of $g(r)$ we have not used to calculate $\chi^2$ since it only captures nearest neighbor features, not the pore structure.
The main difference between the bad fit and the others is the scaling which is due to slow convergence since the bad fit dataset has very large pores.
There are two weak bumps at \SI{6.5}{\angstrom} and \SI{7.5}{\angstrom} which we don't pick up with our model, but the general behaviour is reproduced by our model both on short and long length scales.}
\label{fig:gofr1}
\end{figure}

Finally, we depict a comparison of the actual simulated input SiO$_2$ data set together with a
representation of the best-fit parameters in figure \ref{fig:porous_vs_model}   
\begin{figure}
\includegraphics[width=.45\textwidth]{fig9.png}
\caption{Comparison of our best fit model (left) with parameters $\tau=-0.2$, $\psi=0.9$ and $\nu=0.03$ and the simulated SiO$_2$ input data (right).
Although the pore geometry seems to be similar, we see some artifacts, such as single loose atoms in the generated geometry.
To obtain a structure that can be used further, one would have to passivate the system by adding atoms so each atom has the desired coordination number.
Both the porosity and the surface area of the model and the data are found to be similar.}
\label{fig:porous_vs_model}
\end{figure}

@@
